Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-28 workflow/workflow.Rmd
adress sandrine s comments

#bioconductor:
library(scRNAseq)
library(scone)
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
library(GEOquery)
#github
library(slingshot)
#CRAN
library(doParallel)
library(ggplot2)
library(gplots)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(RColorBrewer)
library(digest)
library(rARPACK)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing numer of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; and (3) inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step. [SD: Include DE as 4th step, to match statement in "Conclusion"?]

@@

```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. This figure has been reproduced with kind permission from [@Fletcher2017].'}
knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
```

@@

The scRNA-seq dataset we work with was generated to study the differentitation of HBCs stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017] [SD: Add reference to Slingshot and ZINB-WaVE biorXiv manuscripts.].

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland. See Figure \@ref(fig:stemcelldiff). In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the publicly-available genes x cells matrix of raw counts [SD: Add URL].

@@

Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to a cluster in the orginal publication.

@@

In the original work [@Fletcher2017], cells were clustered into 14 different clusters, where 151 cells were not assigned to a cluster [SD: -2 labels?].

```{r original}
clus.labels <- colData(core)[, "clusterLabels"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```
Note that there is partial confounding [SD: Did you mean nesting?] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression (DE) analysis. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017 ][SD: Ref to ZIFA.].

Here, we perform dimensionality reduction using the Bioconductor R package `zinbwave`, which fits a zero-inflated negative binomial model (ZINB-WaVE) that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects. For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

[SD: Include Fig. 1 from [@Risso2017]?]
As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[SD: Add ref.]. A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

```{r examineCombineMany, fig.cap="Clusters found using the function RSEC from the clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```
The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).

>>>>

#bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
#github
library(scone)
library(slingshot)
#CRAN
library(doParallel)
library(RColorBrewer)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. This figure has been reproduced with kind permission from (Fletcher et al. 2017).'}
knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
```

@@

The scRNA-seq dataset we work with was generated to study the differentitation of HBCs stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017] [@Risso2017], and [@Street2017].

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland. See Figure \@ref(fig:stemcelldiff). In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

@@

Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to a cluster in the original publication.

@@

In the original work [@Fletcher2017], cells were clustered into 14 different clusters, where 151 cells have a cluster label of `-2`, that is they were not assigned to a cluster in the original publication. [SD: -2 labels? FP: yes]

```{r original}
clus.labels <- colData(core)[, "clusterLabels"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```

Note that there is partial nesting [SD: Did you mean nesting? FP: yes] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression (DE) analysis. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].

Here, we perform dimensionality reduction using the Bioconductor R package `zinbwave`, which fits a zero-inflated negative binomial model (ZINB-WaVE) that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects. See a schematic view of the ZINB-WaVE model in Figure \@ref(fig:zinbschema). For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure has been reproduced with kind permission from (Risso et al. 2017)."}
knitr::include_graphics('zinb_schema.pdf')
```

As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005]. A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

```{r examineCombineMany, fig.cap="Clusters found using the function RSEC from the clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).
